**Evaluation of binary classification pipelines and methods for 16S rRNA gene data**

### 2018-12-07 submission/HHMI_Plot.Rmd
HHMI plot initial commit

---
bibliography: references.bib
csl: mbio.csl
fontsize: 11pt
geometry: margin=1.0in
output:
  word_document: default
  pdf_document: 
    includes:
      in_header: header.tex
    keep_tex: yes
---

@@

```{r LoadData, eval=TRUE, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
######################################################################
#----------------- Load OTU table and MetaData -----------------#
######################################################################

# Read in metadata and select only sample Id and diagnosis columns
meta <- read.delim('../data/metadata.tsv', header=T, sep='\t') %>%
  select(sample, dx, Dx_Bin, fit_result)

# Read in OTU table and remove label and numOtus columns
shared <- read.delim('../data/baxter.0.03.subsample.shared', header=T, sep='\t') %>%
   select(-label, -numOtus)

######################################################################
#---Load .tsv data with CV and test AUC data generated in Python ----#
######################################################################

# Read in AUCs table generated from L2 logistic regression model for all samples:
#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
#       FIT is a feature
#       The scaler is MinMax scaler
logit <- read.delim('../data/process/L2_Logistic_Regression.tsv', header=T, sep='\t') %>%
  select(level_1, AUC) %>% 
  rename(Performance = level_1) %>% 
  mutate(model="L2-Logistic Regression")

logit_summary <- logit %>%  
  group_by(Performance) %>% 
  summarise(meanAUC=mean(AUC), std=sd(AUC))

cv_meanAUC <- logit_summary[1,2]
cv_sdAUC <- logit_summary[1,3]
test_meanAUC <- logit_summary[2,2]
test_sdAUC <- logit_summary[2,3]

# Read in AUCs table generated from l1 SVM linear kernel for all samples
#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
#       FIT is a feature
#       The scaler is Standard scaler
l1svm <- read.delim('../data/process/L1_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
  select(level_1, AUC) %>% 
  rename(Performance = level_1) %>% 
  mutate(model="L1-SVM Linear")

l1svm_summary <- l1svm %>%  
  group_by(Performance) %>% 
  summarise(meanAUC=mean(AUC), std=sd(AUC))

# Read in AUCs table generated from l2 SVM linear kernel for all samples
#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
#       FIT is a feature
#       The scaler is Standard scaler
l2svm <- read.delim('../data/process/L2_SVM_Linear_Kernel.tsv', header=T, sep='\t') %>%
  select(level_1, AUC) %>% 
  rename(Performance = level_1) %>% 
  mutate(model="L2-SVM Linear")

l2svm_summary <- l2svm %>%  
  group_by(Performance) %>% 
  summarise(meanAUC=mean(AUC), std=sd(AUC))

# Read in AUCs table generated from  SVM RBF kernel for all samples
#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
#       FIT is a feature
#       The scaler is Standard scaler
svmRBF <- read.delim('../data/process/SVM_RBF.tsv', header=T, sep='\t') %>%
  select(level_1, AUC) %>% 
  rename(Performance = level_1) %>% 
  mutate(model="SVM RBF")

svmRBF_summary <- svmRBF %>%  
  group_by(Performance) %>% 
  summarise(meanAUC=mean(AUC), std=sd(AUC))

# Read in AUCs table generated from xgboost for all samples
#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
#       FIT is a feature
#       The scaler is MinMax (0-1) scaler
xgboost <- read.delim('../data/process/XGBoost.tsv', header=T, sep='\t') %>%
  select(level_1, AUC) %>% 
  rename(Performance = level_1) %>% 
  mutate(model="XGBoost")

xgboost_summary <- xgboost %>%  
  group_by(Performance) %>% 
  summarise(meanAUC=mean(AUC), std=sd(AUC))

# Read in AUCs table generated from random forest for all samples
#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
#       FIT is a feature
#       The scaler is MinMax (0-1) scaler
rf <- read.delim('../data/process/Random_Forest.tsv', header=T, sep='\t') %>%
  select(level_1, AUC) %>% 
  rename(Performance = level_1) %>% 
  mutate(model="Random Forest ")

rf_summary <- rf %>%  
  group_by(Performance) %>% 
  summarise(meanAUC=mean(AUC), std=sd(AUC))
# Read in AUCs table generated from decision tree for all samples
#       Carcinomas + Adenomas are 1 and Normal is 0 for binary.
#       FIT is a feature
#       The scaler is MinMax (0-1) scaler
dt <- read.delim('../data/process/Decision_Tree.tsv', header=T, sep='\t') %>%
  select(level_1, AUC) %>% 
  rename(Performance = level_1) %>% 
  mutate(model="Decision Tree")

dt_summary <- dt %>%  
  group_by(Performance) %>% 
  summarise(meanAUC=mean(AUC), std=sd(AUC))
```

```{python, cache=FALSE, message=FALSE, warning=FALSE, engine.path = '/Library/Frameworks/Python.framework/Versions/3.6/bin/python3'}
import platform
python_version = platform.python_version()
import sklearn
sklearn_module = sklearn.__version__
```
\linenumbers
__Evaluation of machine learning methods that identify colorectal tumors with microbiota-associated biomarkers__

Begüm D. Topçuoğlu, Jenna Wiens, Mack Ruffin, Patrick D. Schloss

As gut microbiome field continues to grow, there is an ever-increasing demand for reproducible machine learning methods to determine associations between the microbiome and a phenotype of interest. Currently, the use of machine learning in microbiome research lacks clarity and consistency over the modeling pipeline (training, validation and testing steps). There is a need for guidance on how to implement good machine learning practices to generate reproducible and robust models.

Recently, there has been growing interest in using machine learning to identify colorectal tumors with microbiota-associated biomarkers. Colorectal cancer is one of the leading cause of death among cancers in the United States. Colonoscopy as a screening tool is effective, however it is invasive and have a low rate of patient adherence. Previous studies have shown that bacterial abundances in the stool can predict colorectal tumors in the colon and can be used as a non-invasive screening tool. However, the prediction performance of these models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9 [@sze_leveraging_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @zackular_human_2014]. The variation in prediction performance is based in part on differences in the study populations, and in part on the differences in modeling pipelines.

In this study, hemoglobin counts and 16S gene sequences in the stool were used to identify colorectal tumors of `r paste(nrow(meta))` patients as advanced tumors or not. Modeling pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and XGBoost binary classification models. The mean AUCs of these models were `r paste(round(test_meanAUC,2))` ± `r paste(round(test_sdAUC,2))`,  `r paste(round(l1svm_summary[2,2],2))` ± `r paste(round(l1svm_summary[2,3],2))`,  `r paste(round(l2svm_summary[2,2],2))` ± `r paste(round(l2svm_summary[2,3],2))`, `r paste(round(svmRBF_summary[2,2],2))` ± `r paste(round(svmRBF_summary[2,3],2))`, `r paste(round(dt_summary[2,2],2))` ± `r paste(round(dt_summary[2,3],2))`, `r paste(round(rf_summary[2,2],2))` ± `r paste(round(rf_summary[2,3],2))`, and `r paste(round(xgboost_summary[2,2],2))` ± `r paste(round(xgboost_summary[2,3],2))`, respectively. Tree-based methods, namely Decision Tree, Random Forest and XGBoost were less susceptible to overfitting and in general had higher sensitivity and specificity for advanced tumors. Aside from evaluating generalization and classification performance of each model, this study established standards for modeling pipelines of microbiome-associated machine learning models.

\newpage

__References__

<div id="refs"></div>

>>>>

---
title: ""
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
---

@@

![Generalization and classification performance of modeling methods. Tukey box-plots show the cross validation and testing AUC values of the L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and XGBoost binary classification models over the 100 different training/test splits.](../results/figures/AUC_comparison.png)
